---
title: "EDA Final Project"
author: "Manali Shelar"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(arm)
library(ggridges)
library(mgcv)
library(dplyr)
library(modelr)
cb_palette = c("#999999", "#E69F00", "#56B4E9", "#009E73",
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Read the project data file:

```{r}
fuel_data <- read.csv("vehicles.csv")
head(fuel_data)
```


```{r}
#View(fuel_data)
```

```{r}
dim(fuel_data)
```



```{r}
# summary(fuel_data)
```

Data Preparation:

```{r}
fuel_data <- fuel_data[!is.na(fuel_data$displ), ]
```



```{r}
fuel_data$decade <- cut(fuel_data$year, breaks = seq(1980, 2030, by = 10), labels = seq(1980, 2020, by = 10))
```

Data Exploration with Basic plots:


Average Fuel Economy over the years

```{r , fig.height=4, fig.width=8}
plot1 <- fuel_data %>% 
  group_by(year) %>% 
  summarise(avgMPG = mean(comb08), .groups = "drop") %>% 
  ggplot(aes(x=year, y = avgMPG)) +
  geom_line(alpha = 0.5, show.legend = FALSE) +
  geom_point(show.legend = FALSE) +
  scale_x_continuous(breaks =  seq( min(fuel_data$year), max(fuel_data$year), by = 4) ) +
  scale_color_viridis_c() + 
  labs(title = "Average fuel economy over the years", y = "Average MPG", x = "Year")
plot1
```

```{r}
ggsave("plot1.png", plot = plot1, width = 6, height = 4)
```



```{r}
# Plotting histogram
plot2 <- ggplot(fuel_data, aes(x = displ)) +
  geom_histogram(binwidth = 0.2, position = "identity", alpha = 0.7) +
  facet_wrap(~ decade, labeller = label_both) +
  labs(title = "Distribution of Engine Displacement by Decade",
       x = "Engine Displacement (liters)",
       y = "Frequency") +
  scale_fill_viridis_d()
plot2
```


```{r}
ggsave("plot2.png", plot = plot2, width = 6, height = 4)
```



Average Fuel Economy over time by fuel type

```{r , fig.height=4, fig.width=11}
plot3 <- fuel_data %>% 
  filter(fuelType1 != "Electricity") %>%
  group_by(year, fuelType1) %>% 
  summarise(avgMPG = mean(comb08), .groups = "drop") %>% 
  ggplot(aes(x=year, y = avgMPG)) +
  geom_line(aes(color = avgMPG), alpha = 0.5, show.legend = FALSE) +
  geom_point(aes(color = avgMPG), show.legend = FALSE) +
  scale_x_continuous(breaks =  seq( min(fuel_data$year), max(fuel_data$year), by = 4) ) +
  scale_color_viridis_c() + 
  facet_wrap(~ fuelType1, labeller = label_both) +
  labs(title = "Average fuel economy over the years", y = "Average MPG", x = "Year") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 7))
plot3
```

```{r}
ggsave("plot3.png", plot = plot3, width = 6, height = 4)
```



Fuel economy by vehicle class:


```{r}
fuel_data <- fuel_data %>%
  mutate(VClass_category = case_when(
    grepl("Cars|Two Seaters", VClass) ~ "Cars",
    grepl("Van", VClass) ~ "Vans",
    grepl("Pickup Trucks", VClass) ~ "Pickup Trucks",
    grepl("Sport Utility Vehicle", VClass) ~ "SUV",
    grepl("Station Wagons", VClass) ~ "Station Wagons",
    TRUE ~ "Other"
  ))
```



```{r, fig.height=5,  fig.width=6}
plot4 <- fuel_data %>% 
  group_by(VClass) %>% 
  mutate(avgMPG = mean(comb08)) %>% ungroup() %>%
  ggplot(aes(x = avgMPG, y = as.factor(VClass_category))) +
  geom_density_ridges(aes(fill = VClass_category), rel_min_height = 0.001, show.legend = FALSE, alpha = 0.7) +
  scale_fill_manual(values = cb_palette) +
  labs(title = "Fuel economy density by Vehicle class", y = "Vehicle Class", x = "Avg MPG")
plot4
```


```{r}
ggsave("plot4.png", plot = plot4, width = 6, height = 4)
```


Fuel Economy and Engine Displacement for different vehicle class (Year- 2023):

```{r}
plot5 <- fuel_data %>% 
  filter(year == 2023) %>%
  ggplot(aes(y=comb08, x = displ)) + 
  geom_point(aes(color = VClass_category, group = VClass_category), position = "jitter",  na.rm = TRUE, size = 2, alpha = 0.6) +
  scale_fill_manual(values = cb_palette) +
  labs(title = "Fuel Economy vs Displacement", y = "Avg MPG", x = "Displacement (litres)") +
  guides(color = guide_legend(title="Vehicle class"))
plot5
```


```{r}
ggsave("plot5.png", plot = plot5, width = 6, height = 4)
```



Relationship between Fuel economy and Vehicle Class over time:


```{r}
plot6 <- fuel_data %>% 
  group_by(year, VClass_category) %>% 
  mutate(avgMPG = mean(comb08)) %>% 
  ungroup() %>% 
  ggplot(aes(x = year, y = avgMPG)) +
  geom_point(aes(color = VClass_category), alpha = 0.1) +
  geom_smooth(aes(color = VClass_category))  +
  labs(title = "Fuel economy by Vehicle class over time", y = "Avg MPG", x = "Year") +
  guides(color = guide_legend(title="Vehicle class")) +
  scale_fill_manual(values = cb_palette)
plot6
```

```{r}
ggsave("plot6.png", plot = plot6, width = 6, height = 4)
```


Relationship between the Fuel Economy and Displacement Over Time:

```{r warning=FALSE}
ggplot(fuel_data, aes(x = displ, y = comb08, color = year)) +
  geom_point(alpha = 0.2, size = 1) +
  # geom_jitter(height = 0.1, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +  
  labs(title = "Fuel Economy and Displacement Over Time",
       x = "Displacement",
       y = "Combined MPG") +
  theme_minimal()
```



```{r}
plot7 <- ggplot(fuel_data, aes(x = displ, y = comb08)) +
  geom_point(alpha = 0.2, size = 0.6) +
  geom_smooth(method = "gam", se = FALSE) +  
  facet_wrap(~ decade, labeller = label_both) +
  labs(title = "Fuel Economy and Displacement Over Time",
       x = "Displacement",
       y = "Combined MPG") +
  theme_minimal()
plot7
```


```{r}
ggsave("plot7.png", plot = plot7, width = 6, height = 4)
```

One predictor models:

GAM Model:


```{r}
fuel_data.model1 <- gam(comb08 ~ s(displ), data = fuel_data)
fuel_data.model1
```

```{r}
fuel_data.model1.df <- augment(fuel_data.model1)
```

```{r}
ggplot(fuel_data.model1.df, aes(x = displ, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Displacement", y = "Fitted Values")
```



```{r}
ggplot(fuel_data.model1.df, aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() +
  labs(x = "Displacement", y = "Residuals")
```




Two predictor models:

GAM Model:

```{r}
fuel_data.model3 <- gam(comb08 ~ s(displ) + s(year), data = fuel_data)
fuel_data.model3
```

```{r}
plot(fuel_data.model3)
```



```{r}
fuel_data.model3.df <- augment(fuel_data.model3)
```

```{r}
ggplot(fuel_data.model3.df, aes(x = displ, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Displacement", y = "Fitted Values")
```


```{r}
ggplot(fuel_data.model3.df, aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() +
  labs(x = "Displacement", y = "Residuals")
```

Three predictor model:

GAM Model:

```{r}
fuel_data.model5 <- gam(comb08 ~ s(displ) + s(year) + VClass_category, data = fuel_data)
fuel_data.model5

plot(fuel_data.model5)
```






```{r}
fuel_data.model5.df <- augment(fuel_data.model5)
```


```{r}
plot8 <- ggplot(fuel_data.model5.df, aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Displacement", y = "Residuals")
plot8 

ggplot(fuel_data.model5.df, aes(x = year, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Year", y = "Residuals")
ggplot(fuel_data.model5.df, aes(x = VClass_category, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Vehicle Category", y = "Residuals")

plot9 <- ggplot(fuel_data.model5.df %>% 
         mutate(decade = as.factor(floor(year / 10) * 10)), 
       aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() +
  labs(title = "Residual Analysis by Decade", x = "Displacement", y = "Residuals") +
  facet_wrap(~ decade, labeller = label_both)
plot9

ggplot(fuel_data.model5.df, aes(x = .fitted, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Fitted Values", y = "Residuals")
```

```{r}
ggsave("plot8.png", plot = plot8, width = 6, height = 4)
ggsave("plot9.png", plot = plot9, width = 6, height = 4)
```

```{r}
ggplot(fuel_data.model5.df, aes(x = displ, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Displacement", y = "Fitted Values")

ggplot(fuel_data.model5.df, aes(x = year, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Year", y = "Fitted Values")
```


```{r}
fuel.grid2 = expand.grid(displ = seq(0.6, 7.0, 0.1),
                         year = seq(1984, 2024, 1),
                         VClass_category = c('Cars', 'Vans', 'Pickup Trucks', 'SUV', 'Station Wagons'))
mpg.predict2 <- predict(fuel_data.model5, newdata = fuel.grid2)
mpg.df2 <- data.frame(fuel.grid2, comb08 = as.vector(mpg.predict2))
```




```{r , fig.height=4,  fig.width=8}
plot10 <- ggplot(mpg.df2, aes(x = displ, y = comb08, color = year, group = year)) +
  facet_grid(~VClass_category) + geom_line() +
  scale_color_viridis_c() +
  labs(x = "Displacement", y = "Combined MPG", color = "Year")
plot10
```




```{r}
ggsave("plot10.png", plot = plot10, width = 6, height = 4)
```


```{r , fig.height=4,  fig.width=8}
ggplot(mpg.df2 %>% 
         mutate(decade = as.factor(floor(year / 10) * 10)), 
       aes(x = displ, y = comb08, color = VClass_category, group = VClass_category)) +
  labs(x = "Displacement", y = "Residuals") +
  facet_grid(~ decade, labeller = label_both) + geom_line()+
  scale_color_manual(values = cb_palette)
```


Three predictor model with interaction:


```{r}
fuel_data.model6 <- gam(comb08 ~ s(displ, year) + VClass_category, data = fuel_data)
fuel_data.model6

plot(fuel_data.model6)
```



```{r}
fuel_data.model6.df <- augment(fuel_data.model6)
```


```{r}
ggplot(fuel_data.model6.df, aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Displacement", y = "Residuals")
ggplot(fuel_data.model6.df, aes(x = year, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Year", y = "Residuals")
ggplot(fuel_data.model6.df, aes(x = VClass_category, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Vehicle Category", y = "Residuals")

ggplot(fuel_data.model6.df %>% 
         mutate(decade = as.factor(floor(year / 10) * 10)), 
       aes(x = displ, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() +
  labs(x = "Displacement", y = "Residuals") +
  facet_wrap(~ decade, labeller = label_both)

ggplot(fuel_data.model6.df, aes(x = .fitted, y = .resid)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.2, size = 0.6) +
  geom_smooth() + labs(x = "Fitted Values", y = "Residuals")
```

```{r}
ggplot(fuel_data.model6.df, aes(x = displ, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Displacement", y = "Fitted Values")

ggplot(fuel_data.model6.df, aes(x = year, y = .fitted)) +
  geom_jitter(height = 0.25, width = 0.1, alpha = 0.5, size = 0.4) +
  geom_smooth() +
  labs(x = "Year", y = "Fitted Values")
```


```{r}
fuel.grid2 = expand.grid(displ = seq(0.6, 8.4, 0.1),
                         year = seq(1984, 2024, 1),
                         VClass_category = c('Cars', 'Vans', 'Pickup Trucks', 'SUV', 'Station Wagons'))
mpg.predict2 <- predict(fuel_data.model6, newdata = fuel.grid2)
mpg.df2 <- data.frame(fuel.grid2, comb08 = as.vector(mpg.predict2))
```


```{r , fig.height=4,  fig.width=8}
ggplot(mpg.df2, aes(x = displ, y = comb08, color = year, group = year)) +
  facet_grid(~VClass_category) + geom_line() +
  scale_color_viridis_c()
```
















































